---
title: "Gambling Industry Statistics Overview"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---


```{r setup, include=FALSE}

library(flexdashboard)
library(readxl)
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
library(plotly)
library(shiny)
library(billboarder)
library(reshape2)
library(corrplot)


#Import and Clean Data

# Read data from the specified worksheet into a data frame
#file path
file_path="C:/Users/Nacim/OneDrive/Documents/Bioinformatics job preparation/Gambling-Industry-Overview/Data Analyst Exercise.xlsx"
#"C:\Users\Nacim\OneDrive\Documents\Bioinformatics job preparation\Gambling-Industry-Overview\Data Analyst Exercise.xlsx"
Gambling_df <- read_excel(file_path, sheet = "Gambling industry data")
                          
                               #Initial Checks#
#check dimensions
dim(Gambling_df)
#check Structure of dataframe
str(Gambling_df)
# Check the top 10 rows of the dataframe
head(Gambling_df, 10)
# Check the bottom 10 rows of the dataframe
tail(Gambling_df, 10)

#View rows with any missing values
Gambling_df %>% 
filter(!complete.cases(.))

# Check columns classes
sapply(Gambling_df, class)
                               #data cleaning#

#Remove rows with all empty values 
#Use the 7th row as new column names
## Replace spaces in column names with underscores and convert column names to lowercase

Gambling_df <- Gambling_df[!apply(is.na(Gambling_df) | Gambling_df == "", 1, all),]%>%
  `colnames<-`(Gambling_df[7, ]) %>% 
{colnames(.) <- gsub(" ", "_", tolower(colnames(.))); .}

# Check the top rows of the dataframe
head(Gambling_df)

#Delete the first 4 rows by range
Gambling_df = Gambling_df %>% slice(-1:-4)
#check column names and change any that need changing 
colnames(Gambling_df)
Gambling_df <- Gambling_df %>%
rename(`percentage_change` = "%_\r\nchange")

#how many NA left?
#View rows with any missing values
Gambling_df %>% 
filter(!complete.cases(.))

#Change missing value to Zero as it is genuinely missing 
Gambling_df <- Gambling_df %>% 
  mutate(`percentage_change` = as.numeric(`percentage_change`),  # Convert column to numeric
    `percentage_change` = replace_na(`percentage_change`, 0))


#set all data as numeral values instead of character except reporting period column and round to 2 decimal place
Gambling_df <- Gambling_df %>%
  mutate_at(vars(-'reporting_period'), ~ round(as.numeric(.), 2))

#check no duplicates in the reporting period column
any_Duplicate = Gambling_df[duplicated(Gambling_df), ]
print(any_Duplicate)

## Convert the column to a proper date format#
# Split the Reporting Period into start and end dates while keeping original column
# covert new columns to date and add %d
gambling_analysis <- Gambling_df %>%
  separate(`reporting_period`, into = c("start_date", "end_date"), sep = " -", remove = FALSE) %>%
  mutate(start_date = dmy(paste('01', start_date)),
         end_date = dmy(paste('31', end_date))) 

                                    #Finale Data check
#Structure of dataframe
str(gambling_analysis)

#check for missing values
na_count <- colSums(is.na(gambling_analysis)) 

#create summary table
g_summary = data.frame(summary(gambling_analysis))

```

<span style="font-size: small;"> Version 1.0 </span>


General Overview 
=========================================================


Column {.tabset}
-----------------------------------------------------------------------

### Overall Gross Gambling Yield

```{r}
#create function for line graph
create_ggplot <- function(data) {
  ggplot(data, aes(x = start_date, y = overall_total)) +
  geom_line(color = "#0099f9", size = 1) +
  geom_point(color = "#0099f9", size = 3) +
  scale_y_continuous(labels = function(x) paste0(x, "M")) +  # Add "M" to y-axis labels
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Set x-axis labels to display every year
  labs(title = "Great Britain Gross Gambling Yield ",  
       subtitle = "From 2008 to 2022",  
       caption = "Source: Gambling Commission",  
       x = "Reporting Start Date",  
       y = "Total Gross Gambling Yield in GBP") +  
  theme(
    plot.title = element_text(color = "#0099f9", size = 20, face = "bold", hjust = 0.5),  # Customize plot title appearance
    plot.subtitle = element_text(size = 10, face = "bold", hjust = 0.5),  # Customize subtitle appearance
    plot.caption = element_text(face = "italic", hjust = 0),  # Customize caption appearance
  )
}

gg = create_ggplot(gambling_analysis)
# Convert ggplot object to plotly for interactivity
ggplotly(gg)
```

### Trend 

```{r}
############################### trend analysis  using linear regression to model the relationship between time and  the gross gambling yield)
# Fit linear regression model
model <- lm(overall_total ~ start_date, data = gambling_analysis)
regression <- summary(model)

# Create scatter plot with trend line
ggplot(data = gambling_analysis, aes(x = start_date, y = overall_total)) +
  geom_point(color = "#0099f9", size = 3) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Trend Analysis of Gross Gambling Yield",
       x = "Reporting Start Date", y = "Total Gross Gambling Yield") +
  theme_minimal()


```

### Yearly Growth Rate
```{r}
gg <- 
  ggplot(gambling_analysis, aes(x = start_date, y = percentage_change)) +
  geom_line(color = "#006400", size = 1) +
  geom_point(color = "#006400", size = 3) +
  scale_y_continuous(limits = c(-0.2, 0.2), breaks = seq(-0.2, 0.2, by = 0.05)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Set x-axis labels to display every year
  labs(title = "Percentage Change in Great Britain Gross Gambling Yield",  
       subtitle = "From 2008 to 2022",  
       caption = "Source: Gambling Commission",  
       x = "Reporting Start Date",  
       y = "Total Yeild % Change") +  
  theme(
    plot.title = element_text(color = "#006400", size = 15, face = "bold", hjust = 0.5),  # Customize plot title appearance
    plot.subtitle = element_text(size = 10, face = "bold", hjust = 0.5),  # Customize subtitle appearance
    plot.caption = element_text(face = "italic", hjust = 0),  # Customize caption appearance
    panel.grid.minor = element_line(color = "white", linetype = "dotted")  # Add minor grid lines
  )

# Convert ggplot object to plotly for interactivity
ggplotly(gg)

```

Column {data-width=350}
-----------------------------------------------------------------------

### Remote vs Non Remote

```{r}
sectors_remote <- c(
  "betting_(remote)",
  "bingo_(remote)",
  "casino_(remote)"
)

sectors_non_remote <- c(
  "arcades_(non-remote)",
  "betting_(non-remote)",
  "bingo_(non-remote)",
  "casino_(non-remote)"
)

# Filter relevant columns and calculate percentages for remote and non-remote sectors
remote_non_remote_data <- gambling_analysis %>%
  select(end_date, `total_excluding_national_lottery_&_lotteries`, sectors_remote, sectors_non_remote) %>%
  mutate(
    Remote_Total = `betting_(remote)` + `bingo_(remote)` + `casino_(remote)`,
    Non_Remote_Total = `arcades_(non-remote)` + `betting_(non-remote)` + `bingo_(non-remote)` + `casino_(non-remote)`,
    Remote_Percentage = (Remote_Total / `total_excluding_national_lottery_&_lotteries`) * 100,
    Non_Remote_Percentage = (Non_Remote_Total / `total_excluding_national_lottery_&_lotteries`) * 100
  )

# Plot the Remote vs. Non-Remote Contributions over time
gg <- ggplot(remote_non_remote_data, aes(x = end_date)) +
  geom_line(aes(y = Remote_Percentage, color = "Remote"), size = 1) +
  geom_line(aes(y = Non_Remote_Percentage, color = "Non-Remote"), size = 1) +
  scale_x_date(date_breaks = "2 year", date_labels = "%Y") +  # Set x-axis labels to display every 2 years
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 10)) +
  scale_color_manual(values = c("Remote" = "#006400", "Non-Remote" = "red")) +
  labs(title = "Remote vs. Non-Remote Gambling Contributions 
       (Excluding Lotteries)",
       y = "Percentage of Total Revenue (%)",
       x = "Reporting Period") +
  theme_minimal() +
  theme(
    legend.position = "right",  # Move legend to the right side
    plot.title = element_text(color = "#006400", size = 15, face = "bold", hjust = 0.5),  # Customize plot title appearance
    axis.text.y = element_text(size = 10),  # Adjust font size of y-axis labels
    axis.ticks.y = element_line(size = 0.5),  # Adjust tick size of y-axis
    panel.grid.minor.y = element_blank(),  # Remove minor grid lines
    panel.grid.major.y = element_line(color = "gray80", size = 0.5, linetype = "dashed"),  # Add dashed grid lines
    panel.grid.major.x = element_line(color = "gray80", size = 0.5, linetype = "dashed"),  # Add dashed grid lines
  ) 


# Convert ggplot object to plotly for interactivity
gg_plotly <- ggplotly(gg) %>% 
  layout(
    title = list(text = "Percentage Change in Great Britain Gross Gambling Yield", font = list(size = 12)),
     legend = list(
      title = list(text = "Colour", font = list(size = 10)),  # Adjust legend title size
      font = list(size = 8)# Adjust legend label size
    )
  )

# Display the interactive plot
gg_plotly
```

### Summary

A statistically significant positive trend in the gross gambling yield over time suggests that the gambling industry is experiencing consistent growth. The year 2020 marked a period of significant change for the gambling industry, potentially influenced by the effects of COVID-19 and lockdown measures.

There has been a noticeable transition from in-person to remote interactions within the gambling industry. This shift could be attributed to technological advancements, shifts in customer behavior, and the increased accessibility accelerated by response to COVID-19 and lockdowns.

These findings imply that there is a need to update remote licensing and online safety regulations. This is crucial to ensure the protection of individuals and the broader public, especially in light of the evolving landscape shaped by remote gambling trends and the potential risks associated with them.


Sector Analysis
========================================================

Column {.tabset}
-----------------------------------------------------------------------
### Average Gambling Yield  
```{r}
# Calculate average Gross Gambling Yield by sector
average_yield_by_sector <- gambling_analysis %>%
  summarise(across(`arcades_(non-remote)`:`the_national_lottery_(remote_and_non-remote)`, ~mean(.x, na.rm = TRUE)))

# Convert the data to a format suitable for plotting
average_yield_melted <- melt(average_yield_by_sector) %>% 
   rename(sector = variable, average_yield = value)

gg <- ggplot(data = average_yield_melted, aes(x = sector, y = average_yield)) +
  geom_bar(stat = "identity", position = "dodge", fill = "steelblue") +
  labs(title = "Average Gross Gambling Yield by Sector",
       x = "Sector", y = "Average Yield in Million",
       subtitle = "Between 2008 to 2022",
       caption = "Source: Gambling Commission") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# Convert ggplot object to plotly for interactivity
ggplotly(gg) 
  



```

### Gross Gambling Yield Over Time 
```{r}
# Create a ggplot object
gg <- ggplot(data = gambling_analysis, aes(x = start_date)) +
  geom_line(aes(y = `arcades_(non-remote)`, color = "arcades")) +
  geom_line(aes(y = `betting_(non-remote)`, color = "Betting (non-remote)")) +
  geom_line(aes(y = `bingo_(non-remote)`, color = "Bingo (non-remote)")) +
  geom_line(aes(y = `casino_(non-remote)`, color = "Casino (non-remote)")) +
  geom_line(aes(y = `betting_(remote)`, color = "Betting (remote)")) +
  geom_line(aes(y = `bingo_(remote)`, color = "Bingo (remote)")) +
  geom_line(aes(y = `casino_(remote)`, color = "Casino (remote)")) +
  geom_line(aes(y = `lotteries_(remote_and_non-remote)`, color = "Lotteries (remote and non-remote)")) +
  geom_line(aes(y = `the_national_lottery_(remote_and_non-remote)`, color = "The National Lottery (remote and non-remote)")) +
  scale_x_date(date_breaks = "2 year", date_labels = "%Y") +  # Set x-axis labels to display every
  labs(title = "Gross Gambling Yield Over Time by Sector",
       x = "Reporting Period",
       y = "Gross Gambling Yield in Million GBP",
       color = "Sector") +
  theme_minimal() +
  scale_color_manual(values = c("Arcades" = "blue",
                                "Betting (non-remote)" = "green",
                                "Bingo (non-remote)" = "red",
                                "Casino (non-remote)" = "purple",
                                "Betting (remote)" = "orange",
                                "Bingo (remote)" = "pink",
                                "Casino (remote)" = "brown",
                                "Lotteries (remote and non-remote)" = "gray",
                                "The National Lottery (remote and non-remote)" = "black"))



# Convert ggplot object to plotly for interactivity
gg_plotly <- ggplotly(gg) %>% 
  layout(
     legend = list(
      title = list(text = "Sector", font = list(size = 10)),  # Adjust legend title size
      font = list(size = 8),# Adjust legend label size
    x = 0.6,  # Set x position to center
     y = -0.65,    # Set y position at the top
      xanchor = "center",  # Anchor x position at the center
      yanchor = "bottom",  # Anchor y position at the top
    orientation = "h"  # Display legend entries horizontally
    )
  )

# Display the interactive plot
gg_plotly



```



Column {.tabset}
-----------------------------------------------------------------------
### Sector Contributions to Gross Gambling Yeild over time 
```{r}
#Calculate sector contributions for each reporting period
sector_contributions <- gambling_analysis %>% 
  select(`start_date`, `overall_total`,`arcades_(non-remote)`:`the_national_lottery_(remote_and_non-remote)`) %>% #Select relevant columns for calculation
  mutate_at(vars(`arcades_(non-remote)`:`the_national_lottery_(remote_and_non-remote)`), 
            ~ . / overall_total * 100) %>% #Calculate the contribution percentage of each sector
  gather(key = "Sector", value = "Contribution", -`start_date`) %>% # Convert data to long format using gather
  
  filter(Sector != "overall_total") %>% #Filter out rows where the Sector is "overall_total"
  
  arrange(start_date) #Arrange the data by start_date


# Create a stacked area chart to visualize sector contributions over time
gg <- ggplot(data = sector_contributions, aes(x = start_date, y = Contribution, fill = Sector)) +
  geom_bar(position='stack', stat='identity') +
  scale_y_continuous(limits = c(0, 105), breaks = seq(0, 105, by = 10)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Set x-axis labels to display every
  
  labs(title = "Sector Contributions to Overall Gross Gambling Yield",
       x = "Reporting Period", y = "Percentage Contribution (%)",
       fill = "Sector") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.key.size = unit(0.2, "cm"))  # Adjust the size

# Convert ggplot object to plotly for interactivity
gg_plotly <- ggplotly(gg) %>% 
  layout(
    title = list(text = "Sector Contributions to Overall Gross Gambling Yield", font = list(size = 15)),
     legend = list(
      title = list(text = "Sector", font = list(size = 10)),  # Adjust legend title size
      font = list(size = 8),# Adjust legend label size
    x = 0.6,  # Set x position to center
     y = -0.65,    # Set y position at the top
      xanchor = "center",  # Anchor x position at the center
      yanchor = "bottom",  # Anchor y position at the top
    orientation = "h"  # Display legend entries horizontally
    )
  )

# Display the interactive plot
gg_plotly

```

### Apr 2008 - Mar 2009
```{r}

# list sectors
sectors <- c(
  "arcades_(non-remote)",
  "betting_(non-remote)",
  "bingo_(non-remote)",
  "casino_(non-remote)",
  "betting_(remote)",
  "bingo_(remote)",
  "casino_(remote)",
  "lotteries_(remote_and_non-remote)",
  "the_national_lottery_(remote_and_non-remote)"
)

generate_interactive_pie_chart <- function(data, selected_period, sectors_col, overall_total_col) {
  # Filter data for the selected reporting period
  single_period_data <- data %>%
    filter(reporting_period == selected_period) %>%
    select(all_of(sectors_col), all_of(overall_total_col))
  
  # Calculate percentages for each sector based on Overall_Total
  sector_percentages <- single_period_data %>%
    mutate(across(all_of(sectors_col), ~ (. / !!sym(overall_total_col)) * 100)) %>%
    select(-all_of(overall_total_col))
  
  # Reshape data for plotting
  sector_percentages_long <- sector_percentages %>%
    pivot_longer(cols = all_of(sectors_col), names_to = "Sector", values_to = "Percentage")
  
  # Generate the interactive pie chart using billboarder and bb_piechart
  billboarder() %>% 
    bb_piechart(
      data = sector_percentages_long,
      x = "Sector",
      cols = "Percentage",
      expand = TRUE,
      title = paste("Sector Contributions to Overall Total -", selected_period),
  )
}

# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2008 - Mar 2009", sectors, "overall_total")

```

### Apr 2009 - Mar 2010
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2009 - Mar 2010", sectors, "overall_total")
```

### Apr 2010 - Mar 2011
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2010 - Mar 2011", sectors, "overall_total")
```
### Apr 2011 - Mar 2012
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2011 - Mar 2012", sectors, "overall_total")
```

### Apr 2012 - Mar 2013
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2012 - Mar 2013", sectors, "overall_total")
```

### Apr 2013 - Mar 2014
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2013 - Mar 2014", sectors, "overall_total")
```

### Apr 2014 - Mar 2015
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2014 - Mar 2015", sectors, "overall_total")

```
### Apr 2015 - Mar 2016
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2015 - Mar 2016", sectors, "overall_total")

```

### Apr 2016 - Mar 2017
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2016 - Mar 2017", sectors, "overall_total")
  
```

### Apr 2017 - Mar 2018
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2017 - Mar 2018", sectors, "overall_total")
```

### Apr 2018 - Mar 2019
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2008 - Mar 2009", sectors, "overall_total")

```

### Apr 2019 - Mar 2020
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2019 - Mar 2020", sectors, "overall_total")
```

### Apr 2020 - Mar 2021
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2020 - Mar 2021", sectors, "overall_total")
```

### Apr 2021 - Mar 2022
```{r}
# Call the function with your specific data and columns
generate_interactive_pie_chart(gambling_analysis, "Apr 2021 - Mar 2022", sectors, "overall_total")
```

<span style="font-size: small;"> Source: Gambling Commission </span>




